Black swans in ecological time series\\Black swans in population dynamics\\Heavy-tailed ecological time series\\Black swans in ecological dynamics

### 2014-10-17 ms.tex
Edit ms

Here we develop a probabilist framework for identifying ecological black swans in population dynamic process noise --- the stochastic jumps from time step to time step. Our framework quantifies both the magnitude and probability of heavy-tailed dynamics, allows for a range of population dynamic models, can incorporate observation uncertainty, and is applicable to the lengths of abundance time series most commonly available. We apply our model to hundreds of populations from around the world across four taxonomic classes to address three questions: (1) how frequent and strong are black swans in population dynamics across taxonomy, (2) what are some verified causes of black swans in ecological time series, and (3) are there characteristics of time series or intrinsic life-history characteristics associated with heavy-tailed dynamics. We find a high probability of ecological black swans across a wide range of taxa but in a limited number of populations. When they occurred, extreme climate, predation, parasites, and their interactions were common documented causes. We found little evidence that intrinsic life-history characteristics were related to the probability of black swans.

@@

To obtain estimates of the probability and magnitude of black swans, we fitted population dynamic models to hundreds of abundance time series from around the world. The key addition of our models, compared to typical population models, was estimating the shape of the process noise tails --- how extreme the largest jumps from time step to time step were. We then compared how these estimates related to taxonomy, time series properties, and life-history characteristics, and for a subset of populations, we documented the established causes of ecological black swans.

@@

We obtained abundance time series from the Global Population Dynamics Database (GPDD) \citep{gpdd2010}. The GPDD contains nearly 5000 time series of abundance across $\sim$ 1000 species and 100 taxonomic orders. To derive a high-quality subset of populations that were suitable for our analyses, we filtered the data (Supporting Material). Our filtering procedure removed populations from less reliable data sources, removed those without sufficient data for our models, and interpolated some missing values. Our final dataset contained \NPops~populations across \NOrders~taxonomic orders and seven taxonomic classes with a median of XX time steps (range of XX--XX) (Table S1, Fig.~\ref{fig:all-ts}).

@@

Our main analysis focusses on the commonly applied phenomenological Gompertz population dynamics model. In previous analyses of the GPDD, the Gompertz has proved to be the most frequently parsimonious model \citep{brook2006}. We extended the traditional Gompertz model by allowing the process noise to be drawn from a Student-t distribution with degrees of freedom parameter $\nu$. If $\nu$ is small ($\lesssim 10$) the t distribution has much heavier tails than a normal distribution. For example, at $\nu = 2$, the probability of drawing a value more than five standard deviations below the mean is $1.8$\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity the distribution approaches the normal distribution (Fig.~\ref{fig:didactic}). By estimating the value of $\nu$ we can quantity how heavy-tailed the process noise deviations are.

@@

We estimated all models in a Bayesian framework using Stan \citep{stan-manual2014}. Stan samples from the posterior distribution with a version of Hamiltonian Markov chain Monte Carlo called the No-U-Turn Sampler \citep{hoffman2014}. We assured that chains had sufficiently converged and the sampler had obtained sufficient independent samples from the posterior ($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Material). We used simulated data to test how how easily we could detect $\nu$ given different sample sizes and to ensure we could recover unbiased parameter estimates from the Gompertz model (Supporting Material, Fig.~\ref{fig:sim-nu}, \ref{fig:sim-gompertz}, \ref{fig:sim-gompertz-boxplots}).

@@

We found strong evidence that the multiplicative process noise jumps from time step to time step were greater than we'd expect under a normal distribution and a Gompertz or Ricker-logistic model in XX--XX\% of populations assessed. We detected black swans more frequently for longer time series and for populations with less typical magnitude of process error, as we might expect. However, we failed to find strong evidence that black swans were associated with density dependence, population growth rate, or lifespan. In cases that we verified, black swans were often a result of the interaction between elements of extreme climate, predators and parasite cycles, and strong changes in human pressures. Our empirical results, sensitivity tests, and simulation tests suggest that estimating the tail shape of process error is a viable method of detecting heavy-tailed population dynamics. The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century.

Our results clarify, but do not disagree with previous related analyses. Previous work with an older version of the GPDD assessed heavy-tailed behaviour but focused on identifying if the lognormal was the most frequently parsimonious distribution of population abundance \citep{halley2002}. For heavy-tailed distributions, \citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the four-parameter Levy stable distribution and found little information criteria support for these distributions in longer time series. However, black swan events are by definition rare, and the majority of time series we studied are short for these purposes. Therefore, we would not expect to observe black swans in a large proportion of populations. Furthermore, by probabilistically quantifying process noise tail behaviour on a continuous spectrum, and allowing for non-stationary time series and density dependence, our analysis allows for a more nuanced description of the evidence of ecological black swans. In an earlier study, \citet{keitt1998} described heavy (power law) tails in breeding bird population abundance. However, this finding was challenged by \citet{allen2001}, who showed that mixing data across species could create falsely generate heavy tails. This work directly feeds into our use of the t distribution to represent heavy tails, since a t distribution can be represented by a mixture of normal distributions (with inverse-gamma-distributed variances).

We might expect to observe black swan dynamics in ecological time series because of unmodelled intrinsic properties of populations or extrinsic forces acting on populations. A t distribution can be formed through a mixture of normal distributions \citep[in which the variances are inverse-gamma distributed,][]{gelman2014}. Therefore, if we miss some underlying mixture of processes we might expect to observe heavy tails \citep{allen2001}. That process might be an aggregation of populations across space, or population diversity, or some intrinsic change in population variability through time. Extrinsic forces could also cause black swan dynamics (REF). These forces could be extreme themselves. For example, extreme climate, predation from or competition with other species experiencing black swans, or sharp changes in human pressure such hunting, fishing, or habitat destruction might cause black swans. Alternatively, the interaction of multiple ``normal'' forces could give rise to black swan ecological dynamics. This could occur if the interaction nature is synergistic \citep[e.g.][]{kirby2009} or even if non-synergistic interactions experience a rare alignment \citep{denny2009}.

There a number of important caveats when considering the generality of our results. First, the GPDD data represent a taxonomically and geographically biased sample of populations. The longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom. Although we expect that we would find strong but rare evidence for black swans for other taxa around the world, the common forces driving those black swans (e.g.~severe winters in Table 1) likely differ. Second, in a large database with disparate datasources, we cannot ensure the quality of every datapoint. Recording mistakes undoubtedly comprise some of the black swans we detected. Conversely, recording mistakes could mask heavy tails if initial observations are discarded or altered as assumed recording errors. Third, the temporal scale of observation and population dynamics vary considerably across populations in the GPDD and this likely influences the detection of heavy tails. For example, if we make frequent observations relative to generation time (e.g.~for many large bodied mammals) we will average across generations and perhaps not observe the tails. Furthermore, if we record population censuses infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and dampen observed black swan dynamics.

Our results suggest a number of important future issues to address regarding ecological black swans. First, given that black swans can and do occur, can we forecast the probability of black swans in space and time? Furthermore, what management policies allow us to detect them quickly after they happen? Secondly, can we isolate the components of ecological dynamics that experience black swans events by moving from phenomenological models such as the Gompertz to mechanistic models, that take into account, for example, recruitment? Thirdly, the only way we may gain further insight into the mechanisms and covariates of ecological black swans may be to focus on specific geographic and taxonomic subsets of data where longer time series are available \citep[e.g.][]{segura2013}. Finally, if a sizeable fraction of populations experience heavy tails in population process noise, what is the impact of allowing for black swans in forecasts of ecological risk? Risk can be defined as the probability of a specific magnitude of undesired event occurring (REF). Therefore, financial analysts are acutely concerned with the shape of distribution tails (REF). It is likely that a comparable focus in ecology would alter our perception of forecasts of, for example, extinction risk.

@@

We thank members of the Earth to Ocean research group for helpful discussions and comments. We are grateful to the contributors and maintainers of the Global Population Dynamics Database and to Compute Canada's WestGrid high-performance computing resources. Three silhouette images were obtained from \texttt{phylopic.org}: rabbit (Sarah Werning), hoverfly (Gareth Monger), and bird (Jean-Raphaël Guillaumin {[}photography{]} and T. Michael Keesey {[}vectorization{]}) under a Creative Commons Attribution 3.0 Unported license. Funding was provided by a Simon Fraser University Graduate Fellowship (SCA), NSERC (NKD, ABC), the Canada Research Chairs Program (NKD), and TAB TODO?.

>>>>

Here we develop a probabilist framework for identifying ecological black swans in population dynamic process noise --- the stochastic jumps from time step to time step. Our framework quantifies both the magnitude and probability of heavy-tailed dynamics, allows for a range of population dynamic models, can incorporate observation uncertainty, and is applicable to commonly available lengths of abundance time series. We apply our model to hundreds of populations from around the world across four taxonomic classes to address three questions: (1) how frequent and strong are black swans in population dynamics across taxonomy, (2) what are some verified causes of black swans in ecological time series, and (3) are there characteristics of time series or intrinsic life-history characteristics associated with heavy-tailed dynamics. We find a high probability of ecological black swans across a wide range of taxa but in a limited number of populations. When they occurred, extreme climate, predation, parasites, and their interactions were common documented causes. We found little evidence that intrinsic life-history characteristics were related to the probability of black swans.

@@

To obtain estimates of the probability and magnitude of black swans, we fitted population dynamic models to hundreds of abundance time series from around the world. The key addition of our models, compared to typical population models, was estimating the shape of the process noise tails. We then compared how these estimates related to taxonomy, time series properties, and life-history characteristics, and for a subset of populations, we documented the established causes of ecological black swans.

@@

We obtained abundance time series from the Global Population Dynamics Database (GPDD) \citep{gpdd2010}. The GPDD contains nearly 5000 time series of abundance across $\sim$1000 species and $\sim$100 taxonomic orders. To derive a high-quality subset of populations that were suitable for our analyses, we filtered the data (Supporting Material). Our filtering removed populations from less reliable data sources, removed those without sufficient data for our models, and interpolated some missing values. Our final dataset contained \NPops~populations across \NOrders~taxonomic orders and seven taxonomic classes with a median of XX time steps (range of XX--XX) (Table S1, Fig.~\ref{fig:all-ts}).

@@

Our main analysis focusses on the commonly applied phenomenological Gompertz population dynamics model. In previous analyses of the GPDD, the Gompertz has proved to be the most frequently parsimonious model \citep{brook2006}. We extended the traditional Gompertz model by allowing the process noise to be drawn from a Student-t distribution with degrees of freedom parameter $\nu$ (Fig.~\ref{fig:didactic}). If $\nu$ is small ($\lesssim 10$) the t distribution has much heavier tails than a normal distribution. For example, at $\nu = 2$, the probability of drawing a value more than five standard deviations below the mean is $1.8$\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity the distribution approaches the normal distribution. By estimating the value of $\nu$ we can quantity how heavy-tailed the process noise deviations are (Fig.~\ref{fig:didactic}).

@@

We estimated all models in a Bayesian framework using Stan \citep{stan-manual2014}. Stan samples from the posterior distribution with a version of Hamiltonian Markov chain Monte Carlo called the No-U-Turn Sampler \citep{hoffman2014}. We assured that chains had sufficiently converged and the sampler had obtained sufficient independent samples from the posterior ($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Material). We used simulated data to test how easily we could detect $\nu$ given different sample sizes and to ensure we could recover unbiased parameter estimates from the Gompertz model (Supporting Material, Fig.~\ref{fig:sim-nu}, \ref{fig:sim-gompertz}, \ref{fig:sim-gompertz-boxplots}).

@@

We found strong evidence that the multiplicative process noise jumps from time step to time step were greater than we would expect under a normal distribution and a Gompertz or Ricker-logistic model in XX--XX\% of populations assessed. We detected black swans more frequently for longer time series and for populations with a smaller magnitude of process error, as we might expect. However, we failed to find strong evidence that black swans were associated with density dependence, population growth rate, or lifespan. In cases that we verified, black swans were often a result of the interaction between elements of extreme climate, predators and parasite cycles, and strong changes in human pressures. Our empirical results, sensitivity tests, and simulation tests suggest that estimating the tail shape of process error is a viable method of detecting heavy-tailed population dynamics. The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century (REF).

Our results clarify, but do not disagree with previous related analyses. Previous work with an older version of the GPDD assessed heavy-tailed behaviour but focused on identifying if the lognormal distribution was the most frequently parsimonious model of population abundance \citep{halley2002}. For heavy-tailed distributions, \citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the four-parameter Levy stable distribution and found little information criteria support for these distributions in longer time series. However, black swan events are by definition rare, and the majority of time series in the GPDD are short for these purposes. Therefore, we would not expect to observe black swans in a large proportion of populations. Furthermore, by probabilistically quantifying process noise tail behaviour on a continuous spectrum, and allowing for non-stationary time series and density dependence, our analysis allows for a more nuanced description of the evidence of ecological black swans. In an earlier study, \citet{keitt1998} described heavy (power law) tails in breeding bird population abundance. However, this finding was challenged by \citet{allen2001}, who showed that mixing data across species could create falsely generate heavy tails. This rebutal supports our use of the t distribution to represent heavy tails, since a t distribution can be represented by a mixture of normal distributions \citep[with inverse-gamma-distributed variances,][]{gelman2014}.

We might expect to observe black swan dynamics in ecological time series because of unmodelled intrinsic properties of populations or extrinsic forces acting on populations. Since a t-distribution can be formed by a mixture of normal distributions, we could observe heavy tails if we miss some underlying mixture of intrinsic processes \citep{allen2001}. That process might be an aggregation of populations across space, or population diversity, or some intrinsic change in population variability through time. Extrinsic forces could also cause black swan dynamics (REF). These forces could be extreme themselves. For example, extreme climate, predation from or competition with other species experiencing black swans, or sharp changes in human pressure such hunting, fishing, or habitat destruction might cause black swans (REFS). Alternatively, the interaction of multiple ``normal'' extrinsic forces could give rise to black swan ecological dynamics. This could occur if the interaction nature is synergistic \citep[e.g.][]{kirby2009} or even if non-synergistic interactions experience a rare alignment \citep{denny2009}.

There a number of caveats when considering the generality of our results. First, the GPDD data represent a taxonomically and geographically biased sample of populations. Also, the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom. Although we expect that we would find strong but rare evidence for black swans in other large taxonomic or geographic samples, the common forces driving those black swans (e.g.~severe winters in Table~1) likely differ. Second, in a large database with disparate datasources, we cannot ensure the quality of every datapoint. Recording mistakes undoubtedly comprise some of the black swans we detected. Conversely, recording mistakes could mask heavy tails if initial observations are discarded or altered as assumed recording errors. Third, the temporal scale of observation and population dynamics vary considerably across populations in the GPDD and this likely influences the detection of heavy tails. For example, if we make frequent observations relative to generation time (e.g.~for many large bodied mammals) we will average across generations and perhaps not observe the tails. Furthermore, if we record population censuses infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and dampen observed black swan dynamics.

Our results suggest a number of important future issues to address regarding ecological black swans. First, given that black swans can and do occur, can we forecast the probability of black swans in space and time? Furthermore, what management policies allow us to detect them quickly after they happen? Secondly, can we isolate the components of ecological dynamics that experience black swans events by moving from phenomenological models such as the Gompertz to mechanistic models, that take into account, for example, recruitment? We expect that greater insight into the mechanisms and covariates of ecological black swans may be best obtained through specific geographic and taxonomic subsets of data where longer time series are available \citep[e.g.][]{segura2013}. Finally, if a sizeable fraction of populations experience heavy tails in population process noise, what is the impact of allowing for black swans in forecasts of ecological risk? Financial analysts are acutely concerned with the shape of the tail distribution of financial returns since these impact estimates of risk --- the probability of a specific magnitude of undesired event occurring (REF). It is possible that a comparable focus in ecology would alter our perception of ecological forecasts such as extinction risk.

@@

We thank members of the Earth to Ocean research group for helpful discussions and comments. We are grateful to the contributors and maintainers of the Global Population Dynamics Database and to Compute Canada's WestGrid high-performance computing resources. Three silhouette images were obtained from \texttt{phylopic.org} under a Creative Commons Attribution 3.0 Unported license: rabbit (Sarah Werning), hoverfly (Gareth Monger), bird (Jean-Raphaël Guillaumin {[}photography{]} and T. Michael Keesey {[}vectorization{]}). Funding was provided by a Simon Fraser University Graduate Fellowship (SCA), the Natural Sciences and Engineering Research Council of Canada (NKD, ABC), the Canada Research Chairs Program (NKD), and XX-TODO (TAB).
